EuPPollNet: A European database of plant-pollinator networks

### 2024-08-22 Manuscript/EuPPollNet_Ms.qmd
adressing comments from colleagues

EuPPollNet contains one of the largest sets of plant-pollinator networks collated to date, providing a unique opportunity to examine the prevalence of key structural metrics across networks. For example, despite the large theoretical literature on the meaning of a nested structure in plant-pollinator networks [@bascompte2007; @guimaraes2020], where generalist species interact with all other species but specialists interact preferentially only with generalists, this pattern has only been empirically evaluated with a relatively small number of networks [@bascompte2003; @staniczenko2013; @payrato2019], and is still debated how structural metrics such as connectance and nestedness change across latitudes and bioclimatic regions [@olesen2002; @trojelsgaard2013; @song2017].

@@

The database contains `r study_number` independent published and unpublished studies conducted during the time period `r min_year` - `r max_year` in `r country_number` different countries (**Figure 1a** and **Figure 1b**). The studies differ in sampling effort and methodology, and thus documenting sampling methods and sampling effort is an important feature of EuPPollNet. Most studies took place within a single flowering season (`r format(round(one_year, 2), nsmall = 2)`%), sampled a given location for an average of `r format(round(average_sampling_days, 2), nsmall=2)` days, and documented only diurnal interactions, with `r  most_common_method[[1]]` being the most common sampling method (`r format(round(most_common_method[[3]], 2), nsmall = 2)`%). All the studies documented interactions with Hymenopterans (with `r format(round(all_hymnoptera_percent, 2), nsmall = 2)`% considering all Hymenopterans, `r format(round(bee_percent, 2), nsmall = 2)`% only wild bees and `r format(round(bumblee_percent, 2), nsmall = 2)`% only bumblebees), `r format(round(Diptera_percent_metaweb, 2), nsmall = 2)`% documented interactions with Dipterans (with `r format(round(all_diptera_percent, 2), nsmall = 2)`% considering all Dipterans, `r format(round(syrphid_percent, 2), nsmall = 2)`% only syrphids and `r format(round(other_diptera_percent, 2), nsmall = 2)`% recorded syrphids plus bombylids or tachinid flies), `r format(round(Lepidoptera_percent_metaweb, 2), nsmall = 2)`% with Lepidopterans, and `r format(round(Coleoptera_percent_metaweb, 2), nsmall = 2)`% with Coleopterans. The database includes a total of `r prettyNum(interaction_number, big.mark=",")` distinct interactions. Most of the pollinator species belong to the orders Hymenoptera, Diptera, Lepidoptera and Coleoptera (`r format(round(spp_percent_main_order, 2), nsmall = 2)`%), which account for nearly all interactions in this database (`r format(round(main_orders_percent, 2), nsmall = 2)`%). Species that belong to other orders `r format(round(100-spp_percent_main_order, 2), nsmall = 2)`% are not explored in this study as they conduct a minor fraction of the total interactions (`r format(round(100-main_orders_percent, 2), nsmall = 2)`%). Hymenoptera and Diptera contain the highest number of species comprising each approximately 1,000 species in the database. However, the majority of plant-pollinator interactions are from Hymenoptera species (`r format(round(Hymenoptera_interactions, 2), nsmall = 2)`%; **Figure 1c**). Notably, the western honey bee, _Apis mellifera_, represents `r format(round(apis_percent, 2), nsmall = 2)`% of the total interaction records from the database and an average of `r format(round(apis_average_network, 2), nsmall = 2)`% of the total interactions per network.

```{r echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, out.width="110%", fig.cap= "\\textbf{Figure 1}. (\\textbf{a}) Locations of the studies in EuPPollNet across the European continent, showing the total number of pollinator (i.e., orange heptagon) and plant (i.e., green circles) species per study. The sizes of these shapes are proportional to the respective species counts. For visualization purposes, we have selected only a single location per study. (\\textbf{b}) Number of studies by year in EuPPollNet. (\\textbf{c}) Proportion of species and interactions across the four main pollinator orders in EuPPollNet, excluding interactions from \\textit{Apis mellifera}. The total number of species and interactions is indicated in parentheses at the bottom."}

@@

p2  = ggplot(data2, aes(Year, Counts)) +
geom_col(fill = "gray25", width = 0.9, color="black", linewidth = 0.4) +
theme_light() +
theme(plot.margin=unit(c(0,0,0,0), "mm"),
  axis.title.y = element_text(face = "bold", size = 7, vjust = -2),
  axis.title.x = element_text(face = "bold", size = 8),
  axis.text.y = element_text(size = 6),
  axis.text.x = element_text(size = 6,  angle = 90), 
  strip.text.x = element_text(size=12),
   axis.ticks.x = element_line(color="black"),
  strip.background = element_rect(
  color="black", fill=NA, size=0.5, linetype="solid")) + 
ylab("Number of studies") +
coord_cartesian(expand = F, clip = "off") + 
theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  panel.background = element_blank(),
  panel.border = element_rect(colour = "black", fill=NA, size=1)) +
theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(),
panel.border = element_blank(),
axis.line = element_line(color = 'black')) +
scale_y_continuous(breaks = c(0, 4,8,12), labels = c("0", "4","8", "12"), limits = c(0,12)) +
scale_x_continuous(breaks = seq(min(data2$Year), max(data2$Year), by = 2), limits = c(min(data2$Year),max(data2$Year)))+
ggtitle("(b)") +
theme(plot.title = element_text(size = 10,hjust = -0.3, vjust = -5))

@@

#Plot
p3 = ggplot(data_long, aes(x = Variable, y = Value, fill = Pollinator_order)) + 
geom_bar(stat = "identity", width = 0.7) +
theme_classic() +
scale_fill_viridis(discrete=TRUE, direction = -1, name="Pollinator order") +
xlab(NULL) +
ylab("Percentage") + 
scale_y_continuous(labels = percent_format(), expand = c(0, 0)) +
theme(axis.text = element_text(size=6), 
     axis.title.y = element_text(face = "bold", size = 7, vjust = -2),  
     axis.title.x = element_text(face="bold", size = 7),
     legend.position = "top",
     legend.title = element_text(size = 5, face = "bold"),
     legend.text = element_text(size = 5),
     legend.key.size = unit(3, "mm"),
     legend.margin=margin(0,0,0,0),
     legend.box.margin=margin(-8,-5,-5,-5)) +
guides(fill=guide_legend(nrow=2,byrow=TRUE))+
ggtitle("(c)") +
theme(plot.title = element_text(size = 10,hjust = -0.3, vjust = -6)) +
scale_x_discrete(labels = c("Species\n (2,172)", "Interactions\n (358,049)"))

@@

To assess the completeness of plant and pollinator species in the EuPPollNet database at the European level, we used the aforementioned checklists for plants and pollinators. Specifically for plants, we refined the checklist to include only plants occurring in Europe and excluded taxonomic groups not associated with biotic pollination. We did this by first excluding the families considered to have exclusively a wind pollination mode [see @culley2002], and then by filtering out the genera with wind or non-biotic pollination from families that exhibit both biotic and non-biotic pollination modes. Additionally, we manually included exotic species and added unresolved species names that were not present in the accepted names of the checklist at the current version of usage. For pollinators, we compared only the taxonomic coverage of bees, syrphids and butterflies at the European level, as these are the only taxonomic groups with species checklists available in Europe. In addition, the potential number of pollinator species in Europe was estimated by combining the total number of species of bees, syrphids, butterflies and species from other taxonomic groups. The total number of species from other taxonomic groups was extrapolated by assuming that their coverage within the database is equal to the average coverage of bees, syrphids, and butterflies (mean coverage = `r round(mean_coverage, 2)`; sd = `r round(sd_coverage,2)`).

@@

Europe hosts approximately 5,000 species of pollinators and 25,000 species of plants that benefit from animal pollination. EuPPollNet contains a total of `r prettyNum(pollinator_spp_number, big.mark=",")` pollinators and `r prettyNum(plant_spp_number, big.mark=",")` plant species. The coverage of the main pollinator groups occurring in Europe is `r format(round(bee_coverage, 2), nsmall=2)`% for bees, `r format(round(syrphid_coverage, 2), nsmall=2)`% for syrphids and `r format(round(lepidoptera_coverage, 2), nsmall=2)`% for butterflies (see **Figure S1** for coverage at the family level for bees and butterflies, and at the subfamily level for syrphids). Bees constitute `r format(round(bee_interactions_percent, 2), nsmall=2)`% of the interactions in EuPPollNet, and `r format(round(bee_interactions_non_apis_percent, 2), nsmall=2)`% of the interactions when excluding honey bees. Within the database, `r format(round(bee_genera_percent_coverage, 2), nsmall=2)`% of bee genera have at least one species with interaction records, and the average coverage of species at the bee genus level is `r format(round(average_coverage_bee_genera, 2), nsmall=2)`% (**Figure 2**). The presence or absence of interaction records for bees does not follow a phylogenetic pattern ($\lambda$ = `r format(round(lambda_phylosig_bees, 2), nsmall=2)`; _P_ = `r format(round(p_phylosig_bees, 2), nsmall=2)`). The database coverage of all flowering plant species occurring in Europe is `r format(round(plant_coverage, 2), nsmall=2)`% (**Figure 3**), with an average coverage of `r format(round(mean_plant_fam_cov, 2), nsmall=2)`% at the plant family level. Approximately, half of the plant families have at least one species with interaction records (`r format(round(percent_plant_fam_with_data, 2), nsmall=2)`%), and the presence or absence of interaction data for the different plant species also does not follow a statistically relevant phylogenetic pattern ($\lambda$ = `r format(round(lambda_phylosig_plants, 2), nsmall=2)`; _P_ = `r format(round(p_phylosig_plants, 2), nsmall=2)`).

@@

```{r echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, out.width="95%", fig.cap= "\\textbf{Figure 4}. Graphs (a-b-c) indicate the accumulation curves for pollinators, plants, and their interactions across networks. Grey solid lines represent 100 randomized accumulation curves, the black solid lines represent the interpolated curve (i.e., the mean across curves), and the red dashed lines illustrate the extrapolated curve for approximately 3000 networks. The solid black points indicate the number of species and interactions contained in the database. Graph (d) shows the accumulation curve of pollinator species across an increasing number of plant species. This last graph uses the same color and shape structure as the ones in the top panel. Graphs (e-f) indicate the percentage of occurrence (i.e., incidence) of plant and pollinator species across networks. Species on the left (i.e., common) are found in many networks, while species on the right (i.e., rare) are found in few or only a single network. Note that indeed \\textit{Apis mellifera} is the most common pollinator but was excluded from this visualization.", fig.width=10, fig.height=6, out.width="115%"}

@@

#Plot
#Order levels
d_int1_lines$col = factor(d_int1_lines$col, levels = c("collectors", "interpolated", "extrapolated"))
#Plot
p3 = ggplot(d_int1_lines, aes(x = t, y = qD, group = method, color = col)) +
geom_line(aes(linetype = col), size = 0.65) +
scale_colour_manual(values = c("gray", "black", "red"), labels = c("Randomizations", "Interpolated", "Extrapolated")) +
scale_linetype_manual(values = c("solid", "solid", "dashed"),  labels = c("Randomizations", "Interpolated", "Extrapolated")) + 
geom_ribbon(aes(ymin = qD.LCL, 
    ymax=qD.UCL, fill = col),alpha=0.2, colour = NA, show.legend = FALSE)+
scale_fill_manual(values = c("gray", "black", "red")) +
geom_point(data = d_int1_point, col = "black", size=2) + 
ylab("Interactions") +
xlab("Networks") +
theme_bw() +
coord_cartesian(expand = FALSE, clip = "off")  +
xlim(0,3000) +
ylim(0, 30000)+
ggtitle("(c)")

@@

The proportion of species from the major pollinator orders within the database differed across habitats and bioclimatic regions (**Figure 5**). As expected, Hymenoptera was the main taxonomic order on the majority of habitats, exceeded only by Diptera on the habitat categories of riparian vegetation, moors and heathland, and alpine grasslands. Overall, the proportion of Lepidoptera and Coleoptera species was low across all habitats but Coleoptera were notably more abundant in sclerophyllous vegetation and beaches, dunes and sands habitat categories. Similar patterns were observed when exploring the pollinator proportions by bioclimatic region. In this case, Hymenoptera was the predominant order in all bioclimatic regions, with Diptera being particularly abundant in the Alpine and Atlantic regions. Lepidoptera showed low proportions across all bioclimatic regions and Coleoptera were only relevant in the Mediterranean region at European level. Notably, the number of studies (**Figure 5**) and sampling sites (**Figure S2**) also differed across habitats and bioclimatic regions. The habitats sampled by a higher number of studies in the database were `r studies_by_habitat$EuPPollNet_habitat[[1]]` (`r studies_by_habitat$Habitat_count[[1]]`), `r studies_by_habitat$EuPPollNet_habitat[[2]]` (`r studies_by_habitat$Habitat_count[[2]]`) and `r studies_by_habitat$EuPPollNet_habitat[[3]]` (`r studies_by_habitat$Habitat_count[[3]]`). However, the habitats that contain a higher number of sampling sites were `r networks_by_habitat$EuPPollNet_habitat[[1]]` (`r networks_by_habitat$Habitat_count[[1]]`), `r networks_by_habitat$EuPPollNet_habitat[[2]]` (`r networks_by_habitat$Habitat_count[[2]]`) and `r networks_by_habitat$EuPPollNet_habitat[[3]]` (`r networks_by_habitat$Habitat_count[[3]]`). The bioclimatic regions with a higher number of studies were `r str_to_title(studies_by_bioregion$Bioregion[[1]])` (`r studies_by_bioregion$Bioregion_count[[1]]`), `r str_to_title(studies_by_bioregion$Bioregion[[2]])` (`r studies_by_bioregion$Bioregion_count[[2]]`) and `r str_to_title(studies_by_bioregion$Bioregion[[3]])` (`r studies_by_bioregion$Bioregion_count[[3]]`); and those that contain a higher number of sampling sites were `r str_to_title(networks_by_bioregion$Bioregion[[1]])` (`r networks_by_bioregion$Bioregion_count[[1]]`), `r str_to_title(networks_by_bioregion$Bioregion[[2]])` (`r networks_by_bioregion$Bioregion_count[[2]]`) and `r str_to_title(networks_by_bioregion$Bioregion[[3]])` (`r networks_by_bioregion$Bioregion_count[[3]]`).

@@

Connectance values ranged between `r round(min_connectance, 2)` to `r round(max_connectance, 2)` ($\bar{x}$ = `r round(mean_connectance, 2)`) and as expected, followed a negative exponential relationship with the number of species per network (Kendall $\tau$ = `r round(connectance_spp_mean_corr_tau, 2)`, _P_ < 0.01; **Figure 6a**). Nestedness values (NODFc) ranged between `r round(min_nodfc, 2)` to `r round(max_nodfc, 2)` ($\bar{x}$ = `r round(mean_nodfc, 2)`), and as expected for this metric, were independent of the mean number of species (Kendall $\tau$ = `r round(normalised_nestedness_spp_mean_corr_tau, 2)`, _P_ = `r round(normalised_nestedness_spp_mean_corr_pval, 2)`; **Figure S3**). We found that only `r round(over_z_percent, 2)`% of networks were statistically more nested than expected by chance, with `r round(no_diff_z_percent, 2)`% showing no statistical difference, and none being less nested than null expectations (**Figure 6b**). Latitude explained little of the observed variability of residual connectance and nestedness across networks (connectance: R$^2$ = `r round(connectance_lat_r2, 2)`, _P_ = `r round(connectance_lat_pval, 2)`, **Figure 6a**; NODFc: R$^2$ = `r round(normalised_nestedness_lat_r2, 2)`, _P_ = `r round(normalised_nestedness_lat_pval, 2)`, **Figure 6b**). Overall, networks towards higher latitudes showed lower residual connectance and higher nestedness than networks located at lower latitudes. Note that residual connectance and normalised nestedness showed a moderate negative correlation (Kendall $\tau$ = `r round(resid_connectance_nestedness_corr_tau, 2)`, _P_ = `r round(resid_connectance_nestedness_corr_pval, 2)`).

@@

Bees are responsible for the majority of the interactions at the metaweb level, but their relevance for network topology changed across habitats and bioclimatic regions. For instance, plant-pollinator communities in the Mediterranean were dominated by bees, while communities in Alpine or Boreal regions were fly species rich or fly-dominated. These patterns are consistent with our current understanding of bee diversity, which peaks in dry or temperate areas [@orr2021; @leclercq2023]; and with the fact that colder environments (i.e., altitude and latitude wise) harbor a larger fraction of fly pollinators compared to other taxa [@elberling1999; @lefebvre2018]. In addition, beetles were only common floral visitors in the Mediterranean region. Although the networks from the database are visitation networks and do not capture pollinator efficiency [@ballantyne2015], the high proportion of beetles as floral visitors in the Mediterranean provides further support for their potential role as pollinators in this region [@herrera2019; @leon2022]. The number of butterfly species and interactions were relatively low compared to the other taxa. While Europe contains fewer butterfly species than other regions of the world [@ollerton2017], their relevance as pollinators is likely underestimated within this database. This is because a large fraction of studies (~40%) did not sample butterflies, and conventional sampling methods for monitoring other insect pollinators (e.g., bees or flies) may be inadequate for sampling plant-butterfly interactions [@isaac2011]. Honey bees were present in `r round(apis_percent_networks, 1)`% of networks and conducted on average a third of the total interactions per network. The proportion of honey bees in networks across Europe is higher than in natural communities across the world [~13%\; @hung2018], potentially reflecting the dominance of intensive grasslands habitats in EuPPollNet, the highly generalised nature of honeybees, the important role that honeybees are playing as pollinators in Europe, and/or because beekeeping is widely practiced in Europe [@steffan2000; @magrach2017; @herrera2020].

@@

>>>>

EuPPollNet contains one of the largest sets of plant-pollinator networks collated to date, providing a unique opportunity to examine the prevalence of key structural metrics across networks. For example, despite the large theoretical literature on the meaning of a nested structure in plant-pollinator networks [@bascompte2007; @guimaraes2020], where specialists species interact only with subsets of generalists species, this pattern has only been empirically evaluated with a relatively small number of networks [@bascompte2003; @staniczenko2013; @payrato2019], and is still debated how structural metrics such as connectance and nestedness change across latitudes and bioclimatic regions [@olesen2002; @trojelsgaard2013; @song2017].

@@

The database contains `r study_number` independent published and unpublished studies conducted during the time period `r min_year` - `r max_year` in `r country_number` different countries (**Figure 1a** and **Figure 1b**; see **Figure S1** for exact locations). The studies differ in sampling effort and methodology, and thus documenting sampling methods and sampling effort is an important feature of EuPPollNet. Most studies took place within a single flowering season (`r format(round(one_year, 2), nsmall = 2)`%), sampled a given location for an average of `r format(round(average_sampling_days, 2), nsmall=2)` days, and exclusively sampled diurnal plant-pollinator interactions, with `r  most_common_method[[1]]` being the most common sampling method (`r format(round(most_common_method[[3]], 2), nsmall = 2)`%). All the studies documented interactions with Hymenopterans (with `r format(round(all_hymnoptera_percent, 2), nsmall = 2)`% considering all Hymenopterans, `r format(round(bee_percent, 2), nsmall = 2)`% only wild bees and `r format(round(bumblee_percent, 2), nsmall = 2)`% only bumblebees), `r format(round(Diptera_percent_metaweb, 2), nsmall = 2)`% documented interactions with Dipterans (with `r format(round(all_diptera_percent, 2), nsmall = 2)`% considering all Dipterans, `r format(round(syrphid_percent, 2), nsmall = 2)`% only syrphids and `r format(round(other_diptera_percent, 2), nsmall = 2)`% recorded syrphids plus bombylids or tachinid flies), `r format(round(Lepidoptera_percent_metaweb, 2), nsmall = 2)`% with Lepidopterans, and `r format(round(Coleoptera_percent_metaweb, 2), nsmall = 2)`% with Coleopterans. The database includes a total of `r prettyNum(interaction_number, big.mark=",")` distinct interactions. Most of the pollinator species belong to the orders Hymenoptera, Diptera, Lepidoptera and Coleoptera (`r format(round(spp_percent_main_order, 2), nsmall = 2)`%), which account for nearly all interactions in this database (`r format(round(main_orders_percent, 2), nsmall = 2)`%). Species that belong to other orders `r format(round(100-spp_percent_main_order, 2), nsmall = 2)`% are not explored in this study as they conduct a minor fraction of the total interactions (`r format(round(100-main_orders_percent, 2), nsmall = 2)`%). Hymenoptera and Diptera contain the highest number of species comprising each approximately 1,000 species in the database. However, the majority of plant-pollinator interactions are from Hymenoptera species (`r format(round(Hymenoptera_interactions, 2), nsmall = 2)`%; **Figure 1c**). Notably, the western honey bee, _Apis mellifera_, represents `r format(round(apis_percent, 2), nsmall = 2)`% of the total interaction records from the database and an average of `r format(round(apis_average_network, 2), nsmall = 2)`% of the total interactions per network.

```{r echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, out.width="105%", fig.cap= "\\textbf{Figure 1}. (\\textbf{a}) Locations of the studies in EuPPollNet across the European continent, showing the total number of pollinator (i.e., orange heptagon) and plant (i.e., green circles) species per study. The sizes of these shapes are proportional to the respective species counts. For visualization purposes, we have selected only a single location per study. (\\textbf{b}) Number of studies by year in EuPPollNet. (\\textbf{c}) Proportion of species and interactions across the four main pollinator orders in EuPPollNet, excluding interactions from \\textit{Apis mellifera}. The total number of species and interactions is indicated in parentheses at the bottom."}

@@

p2  = ggplot(data2, aes(Year, Counts)) +
geom_col(fill = "gray25", width = 0.9, color="black", linewidth = 0.4) +
theme_light() +
theme(plot.margin=unit(c(0,0,0,0), "mm"),
  axis.title.y = element_text(face = "bold", size = 7, vjust = -2),
  axis.title.x = element_text(face = "bold", size = 8),
  axis.text.y = element_text(size = 6),
  axis.text.x = element_text(size = 6,  angle = 90), 
  strip.text.x = element_text(size=12),
   axis.ticks.x = element_line(color="black"),
  strip.background = element_rect(
  color="black", fill=NA, size=0.5, linetype="solid")) + 
ylab("Number of studies") +
coord_cartesian(expand = F, clip = "off") + 
theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  panel.background = element_blank(),
  panel.border = element_rect(colour = "black", fill=NA, size=1)) +
theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(),
panel.border = element_blank(),
axis.line = element_line(color = 'black')) +
scale_y_continuous(breaks = c(0, 4,8,12), labels = c("0", "4","8", "12"), limits = c(0,12)) +
scale_x_continuous(breaks = seq(min(data2$Year), max(data2$Year), by = 2), limits = c(min(data2$Year),max(data2$Year)))+
ggtitle("(b)") +
theme(plot.title = element_text(size = 10,hjust = -0.4, vjust = -5))

@@

#Plot
p3 = ggplot(data_long, aes(x = Variable, y = Value, fill = Pollinator_order)) + 
geom_bar(stat = "identity", width = 0.7) +
theme_classic() +
scale_fill_viridis(discrete=TRUE, direction = -1, name="Pollinator\n order") +
xlab(NULL) +
ylab("Percentage") + 
scale_y_continuous(labels = percent_format(), expand = c(0, 0)) +
theme(axis.text = element_text(size=6), 
     axis.title.y = element_text(face = "bold", size = 7, vjust = -2),  
     axis.title.x = element_text(face="bold", size = 7),
     legend.position = "top",
     legend.title = element_text(size = 5, face = "bold"),
     legend.text = element_text(size = 5),
     legend.key.size = unit(3, "mm"),
     legend.margin=margin(0,2,0,0),
     legend.box.margin=margin(-9,-9,-6,-6)) +
guides(fill=guide_legend(nrow=2,byrow=TRUE))+
ggtitle("(c)") +
theme(plot.title = element_text(size = 10,hjust = -0.4, vjust = -6)) +
scale_x_discrete(labels = c("Species\n (2,172)", "Interactions\n (358,049)"))

@@

To assess the completeness of plant and pollinator species in the EuPPollNet database at the European level, we used the aforementioned checklists for plants and pollinators. Specifically for plants, we refined the checklist to include only plants occurring in Europe and excluded taxonomic groups not associated with biotic pollination. We did this by first excluding the families considered to have exclusively a wind pollination mode [see @culley2002], and then by filtering out the genera with wind or non-biotic pollination from families that exhibit both biotic and non-biotic pollination modes. Additionally, we manually included exotic species and added unresolved species names that were not present in the accepted names of the checklist at the current version of usage. For pollinators, we evaluated the taxonomic coverage of taxonomic groups with species checklists available in Europe (i.e., bees, syrphids and butterflies). While there is not a good understanding of pollinator diversity in other taxonomic groups, it was assumed that their coverage within the database is equal to the average coverage of bees, syrphids, and butterflies (mean coverage = `r round(mean_coverage, 2)`; sd = `r round(sd_coverage,2)`). Therefore, the total number of of flower visiting species from other taxonomic groups (i.e., non-bee, non syrphid and non-butterfly flower visitors) at European level was extrapolated by assuming that their coverage is equal to the mean coverage of bees, syrphids, and butterflies. Consequently, we provide an estimate for the total number of pollinators across the European continent.

@@

Europe hosts approximately 5,000 species of pollinators and 25,000 species of plants that benefit from animal pollination. EuPPollNet contains a total of `r prettyNum(pollinator_spp_number, big.mark=",")` pollinators and `r prettyNum(plant_spp_number, big.mark=",")` plant species. The coverage of the main pollinator groups occurring in Europe is `r format(round(bee_coverage, 2), nsmall=2)`% for bees, `r format(round(syrphid_coverage, 2), nsmall=2)`% for syrphids and `r format(round(lepidoptera_coverage, 2), nsmall=2)`% for butterflies (see **Figure S2** for coverage at the family level for bees and butterflies, and at the subfamily level for syrphids). Bees (i.e., Anthophila) constitute `r format(round(bee_interactions_percent, 2), nsmall=2)`% of the interactions in EuPPollNet, and `r format(round(bee_interactions_non_apis_percent, 2), nsmall=2)`% of the interactions when excluding honey bees. Within the database, `r format(round(bee_genera_percent_coverage, 2), nsmall=2)`% of bee genera have at least one species with interaction records, and the average coverage of species at the bee genus level is `r format(round(average_coverage_bee_genera, 2), nsmall=2)`% (**Figure 2**). The presence or absence of interaction records for bees does not follow a phylogenetic pattern ($\lambda$ = `r format(round(lambda_phylosig_bees, 2), nsmall=2)`; _P_ = `r format(round(p_phylosig_bees, 2), nsmall=2)`). The database coverage of all flowering plant species occurring in Europe is `r format(round(plant_coverage, 2), nsmall=2)`% (**Figure 3**), with an average coverage of `r format(round(mean_plant_fam_cov, 2), nsmall=2)`% at the plant family level. Approximately, half of the plant families have at least one species with interaction records (`r format(round(percent_plant_fam_with_data, 2), nsmall=2)`%), and the presence or absence of interaction data for the different plant species also does not follow a statistically relevant phylogenetic pattern ($\lambda$ = `r format(round(lambda_phylosig_plants, 2), nsmall=2)`; _P_ = `r format(round(p_phylosig_plants, 2), nsmall=2)`).

@@

```{r echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, out.width="95%", fig.cap= "\\textbf{Figure 4}. Graphs (a-b-c) indicate the accumulation curves for pollinators, plants, and their unique pairwise interactions across networks. Grey solid lines represent 100 randomized accumulation curves, the black solid lines represent the interpolated curve (i.e., the mean across curves), and the red dashed lines illustrate the extrapolated curve for approximately 3000 networks. The solid black points indicate the number of species and interactions contained in the database. Graph (d) shows the accumulation curve of pollinator species across an increasing number of plant species. This last graph uses the same color and shape structure as the ones in the top panel. Graphs (e-f) indicate the percentage of occurrence (i.e., incidence) of plant and pollinator species across networks. Species on the left (i.e., common) are found in many networks, while species on the right (i.e., rare) are found in few or only a single network. Note that \\textit{Apis mellifera} is the most common pollinator but was excluded from this visualization.", fig.width=10, fig.height=6, out.width="115%"}

@@

#Plot
#Order levels
d_int1_lines$col = factor(d_int1_lines$col, levels = c("collectors", "interpolated", "extrapolated"))
#Plot
p3 = ggplot(d_int1_lines, aes(x = t, y = qD, group = method, color = col)) +
geom_line(aes(linetype = col), size = 0.65) +
scale_colour_manual(values = c("gray", "black", "red"), labels = c("Randomizations", "Interpolated", "Extrapolated")) +
scale_linetype_manual(values = c("solid", "solid", "dashed"),  labels = c("Randomizations", "Interpolated", "Extrapolated")) + 
geom_ribbon(aes(ymin = qD.LCL, 
    ymax=qD.UCL, fill = col),alpha=0.2, colour = NA, show.legend = FALSE)+
scale_fill_manual(values = c("gray", "black", "red")) +
geom_point(data = d_int1_point, col = "black", size=2) + 
ylab("Unique interactions") +
xlab("Networks") +
theme_bw() +
coord_cartesian(expand = FALSE, clip = "off")  +
xlim(0,3000) +
ylim(0, 30000)+
ggtitle("(c)")

@@

The proportion of species from the major pollinator orders within the database differed across habitats and bioclimatic regions (**Figure 5**). Hymenoptera was the main taxonomic order in the majority of habitats, exceeded only by Diptera for the habitat categories of riparian vegetation, moors and heathland, and alpine grasslands. Overall, the proportion of flower visitors from Lepidoptera and Coleoptera were low across all habitats but Coleopteran flower visitors were notably more abundant in sclerophyllous vegetation and beaches, dunes and sands habitat categories. Similar patterns were observed when exploring the pollinator proportions by bioclimatic region. Hymenopterans were predominant across all bioclimatic regions and Dipterans were particularly abundant in the Alpine and Atlantic regions. Lepidopterans had low proportions across all bioclimatic regions and Coleopterans were only relevant in the Mediterranean region at European level. Notably, the number of studies (**Figure 5**) and sampling sites (**Figure S3**) also differed across habitats and bioclimatic regions. The habitats sampled by a higher number of studies in the database were `r studies_by_habitat$EuPPollNet_habitat[[1]]` (`r studies_by_habitat$Habitat_count[[1]]`), `r studies_by_habitat$EuPPollNet_habitat[[2]]` (`r studies_by_habitat$Habitat_count[[2]]`) and `r studies_by_habitat$EuPPollNet_habitat[[3]]` (`r studies_by_habitat$Habitat_count[[3]]`). However, the habitats that contain a higher number of sampling sites were `r networks_by_habitat$EuPPollNet_habitat[[1]]` (`r networks_by_habitat$Habitat_count[[1]]`), `r networks_by_habitat$EuPPollNet_habitat[[2]]` (`r networks_by_habitat$Habitat_count[[2]]`) and `r networks_by_habitat$EuPPollNet_habitat[[3]]` (`r networks_by_habitat$Habitat_count[[3]]`). The bioclimatic regions with a higher number of studies were `r str_to_title(studies_by_bioregion$Bioregion[[1]])` (`r studies_by_bioregion$Bioregion_count[[1]]`), `r str_to_title(studies_by_bioregion$Bioregion[[2]])` (`r studies_by_bioregion$Bioregion_count[[2]]`) and `r str_to_title(studies_by_bioregion$Bioregion[[3]])` (`r studies_by_bioregion$Bioregion_count[[3]]`); and those that contain a higher number of sampling sites were `r str_to_title(networks_by_bioregion$Bioregion[[1]])` (`r networks_by_bioregion$Bioregion_count[[1]]`), `r str_to_title(networks_by_bioregion$Bioregion[[2]])` (`r networks_by_bioregion$Bioregion_count[[2]]`) and `r str_to_title(networks_by_bioregion$Bioregion[[3]])` (`r networks_by_bioregion$Bioregion_count[[3]]`).

@@

Connectance values ranged between `r round(min_connectance, 2)` to `r round(max_connectance, 2)` ($\bar{x}$ = `r round(mean_connectance, 2)`) and followed a negative exponential relationship with the number of species per network (Kendall $\tau$ = `r round(connectance_spp_mean_corr_tau, 2)`, _P_ < 0.01; **Figure 6a**). Nestedness values (NODFc) ranged between `r round(min_nodfc, 2)` to `r round(max_nodfc, 2)` ($\bar{x}$ = `r round(mean_nodfc, 2)`), and were independent of the mean number of species (Kendall $\tau$ = `r round(normalised_nestedness_spp_mean_corr_tau, 2)`, _P_ = `r round(normalised_nestedness_spp_mean_corr_pval, 2)`; **Figure S4**). We found that only `r round(over_z_percent, 2)`% of networks were statistically more nested than expected by chance, with `r round(no_diff_z_percent, 2)`% showing no statistical difference, and none being less nested than null expectations (**Figure 6b**). Latitude explained little of the observed variability of residual connectance and nestedness across networks (connectance: R$^2$ = `r round(connectance_lat_r2, 2)`, _P_ = `r round(connectance_lat_pval, 2)`, **Figure 6a**; NODFc: R$^2$ = `r round(normalised_nestedness_lat_r2, 2)`, _P_ = `r round(normalised_nestedness_lat_pval, 2)`, **Figure 6b**). Overall, networks towards higher latitudes showed lower residual connectance and higher nestedness than networks located at lower latitudes. Note that residual connectance and normalised nestedness showed a moderate negative correlation (Kendall $\tau$ = `r round(resid_connectance_nestedness_corr_tau, 2)`, _P_ = `r round(resid_connectance_nestedness_corr_pval, 2)`).

@@

Bees are responsible for the majority of the sampled interactions at the metaweb level. As not all surveys included all pollinator groups, this result may partly be influenced by the taxonomic groups sampled across studies. However, the relevance of bees and other pollinator orders for network topology changed across habitats and bioclimatic regions in accordance to the literature. For instance, plant-pollinator communities in the Mediterranean were dominated by bees, while communities in Alpine or Boreal regions were fly species rich or fly-dominated. These patterns are consistent with our current understanding of bee diversity, which peaks in dry or temperate areas [@orr2021; @leclercq2023]; and with the fact that colder environments (i.e., altitude and latitude wise) harbor a larger fraction of fly pollinators compared to other taxa [@elberling1999; @lefebvre2018]. In addition, beetles were only commonly documented as floral visitors in the Mediterranean region. This study cannot determine whether pollination ecologists traditionally document flower-beetle interactions only in the Mediterranean, or if there are fewer flower visitations by beetles outside this region. Nevertheless, the high proportion of beetles as floral visitors in the Mediterranean provides further support for their potential role as pollinators in this region [@herrera2019; @leon2022]. The number of butterfly species and interactions were relatively low compared to the other taxa. While Europe contains fewer butterfly species than other regions of the world [@ollerton2017], their relevance as pollinators is likely underestimated within this database. This is because a large fraction of studies (~40%) did not sample butterflies, and conventional sampling methods for monitoring other insect pollinators (e.g., bees or flies) may be inadequate for sampling plant-butterfly interactions [@isaac2011]. Honey bees were present in `r round(apis_percent_networks, 1)`% of networks and conducted on average a third of the total interactions per network. The proportion of honey bees in networks across Europe is higher than in natural communities across the world [~13%\; @hung2018], potentially reflecting the dominance of intensive grasslands habitats in EuPPollNet, the highly generalised nature of honeybees, the important role that honeybees are playing as pollinators in Europe, and/or because beekeeping is widely practiced in Europe [@steffan2000; @magrach2017; @herrera2020].

@@

- Figure S4
